Multi-domain, multilingual translation with latent multi-task group dropout

### 2021-11-15 2022/acl/acl_latex.tex
local changes
revised related work and conclusion

\section{Introduction}
Multidomain and multilingual machine translation aim to develop one single model to perform translations for multiple domains and multiple language pairs, respectively.\footnote{We will refer to these two situations as 'multi-task MT' (or \mtsquare{} for short), and refer to individual domains and languages as 'tasks'.} These paradigms are motivated by the compactness of the resulting translation system \citep{Chu18multilingual,dabre20survey}, the hypothetical positive knowledge transfer between similar domains \citep{Pham21revisiting} or between languages in the same family \citep{Tan19multilingual}. However, having all the tasks use exactly the same model parameters can cause negative interference between unrelated tasks \citep{conneau20unsupervised,wang20negative}. Hence the recent development of approaches relying on a partial sharing of the parameters, eg.\ using adapter layers as studied in \citep{houlsby19parameter,Bapna19simple,Pham20Study}. If these techniques have proven effective for building strong baselines, they fail to fully take advantage of the similarities that exist between domains and tasks, as documented eg.\ in \citep{Pham21revisiting}. This is because the partition of the parameter space between generic or task-specific subparts, and their allocation to each task is hard-coded in the network, irrespective of the actual commonalties and differences in the data space.

In this work, we develop techniques that can take the similarity between tasks into account in a more effective way, by learning the network organization from the data. To this end, we introduce a set of latent variables in the model, to account for the unseen association between tasks and regions of the parameter space, and show how training can still be performed end-to-end using a variational surrogate of the standard loss function. Our experiments with multilingual and multidomain machine translation confirm that this method can automatically detect similarities in the data, meaning that related tasks use the same subparts of the network. Our results also show that this method is comparable to using adapter layers in a number of empirical comparisons; however, contrarily to adapters, these performance are obtained without any increase of the model size. Our main contributions are summarized below:
\begin{itemize}
  \setlength{\itemsep}{1pt}
  \setlength{\parskip}{0pt}
  \setlength{\parsep}{0pt}
\item We provide a sound mathematical formulation to the problem of jointly learning task-dependent sub-networks and the parameters of the underlying models using variational probabilistic modeling approach;
\item We present algorithms to train this model end to end with very little extra computational cost.
\item We report, using an extensive experimental setting, significant gains of this method with respect to strong baselines;
\item We study how this method can actually exploit the similarities between tasks to learn interpretable sub-networks.
\end{itemize}
\fyDone{Can you do distillation ? That is only keep the parameter for one domain and still get good results (better than DA ?)}

@@

For multilingual case TED-related, we have four language families including (1) Turkic family with Azerbaijani and Turkish(\domain{aze},\domain{tur}); (2) Slavic family with Belarusian and Russian (\domain{bel},\domain{rus}); (3) Romance family with Glacian and Portuguese (\domain{glg},\domain{por}); and (4) Czech-Slovak family with Slovak and Czech (\domain{ces},\domain{slk}). The similarities between the sub-networks of each pair are 0.95, 0.94, 0.99 and 0.96 which are the maximum in each row except the diagonal, which stands for the self-similarity.
\begin{table*}[h!]
  \centering
  \begin{tabular}{|p{1cm}|*{8}{r|}} \hline
    & \multicolumn{1}{c|}{\domain{aze}} & \multicolumn{1}{c|}{\domain{bel}}& \multicolumn{1}{c|}{\domain{ces}} & \multicolumn{1}{c|}{\domain{glg}} & \multicolumn{1}{c|}{\domain{por}} & \multicolumn{1}{c|}{\domain{rus}} &\multicolumn{1}{c|}{\domain{slk}}&\multicolumn{1}{c|}{\domain{tur}} \\ \hline 
    \domain{aze} &1.0&0.90&0.89&0.86&0.85&0.91&0.89&\SB{0.95}\\
    \domain{bel} &0.90&1.0&0.94&0.91&0.91&\SB{0.94}&0.94&0.88\\
    \domain{ces} &0.89&0.94&1.0&0.92&0.94&0.98&\SB{0.99}&0.91\\
    \domain{glg} &0.86&0.91&0.92&1.0&\SB{0.96}&0.92&0.93&0.86\\ 
    \domain{por} &0.85&0.91&0.94&\SB{0.96}&1.0&0.93&0.94&0.87\\ 
    \domain{rus} &0.91&\SB{0.94}&0.98&0.92&0.93&1.0&0.98&0.93\\
    \domain{slk} &0.89&0.94&\SB{0.99}&0.93&0.94&0.98&1.0&0.91\\
    \domain{tur} &\SB{0.95}&0.88&0.91&0.86&0.87&0.93&0.91&1.0\\
    \hline
  \end{tabular}
  \caption{Multilingual sub-networks' similarities (M2O Ted-related)}
  \label{tab:related-sim}
\end{table*}

We also plot the languages based on their dropouts in Figure\ref{fig:visual}.
\begin{figure*}[htbp]
\begin{subfigure}{\textwidth}
  \centering
  % include first image
  \includegraphics[width=0.7\textwidth]{ted_related}  
  \caption{TED-Related}
  \label{fig:visual-related}
\end{subfigure}
\\
\begin{subfigure}{\textwidth}
  \centering
  % include second image
  \includegraphics[width=0.7\textwidth]{ted_diverse}  
  \caption{TED-Diverse}
  \label{fig:visual-diverse}
\end{subfigure}
\caption{Visualization of languages according to their dropouts (a large vector concatenating the dropping masks of all the layers of the model) constructed by PCA.}
\label{fig:visual}
\end{figure*}

\begin{figure*}[htbp]
\includegraphics[width=0.7\textwidth]{multi_domain}
\caption{Visualization of domains according to their dropouts (a large vector concatenating the dropping masks of all the layers of the model) constructed by PCA.}
\label{fig:domain}
\end{figure*}

\section{Related work}
Multidomain, multilingual translation

\citep{sen19multilingual,kong21multilingual} used the similar idea of language proximity-based parameter sharing. The authors proposed a multi-decoder model sharing the same encoder among languages and routed languages in different families to different decoders.

\textbf{Sparse Transformer}\citet{xian20deep} proposed using different depths for different tasks. The authors aimed to match the depth of the sub-network to the complexity of the task. \citet{Gong21pay,Gong21adaptive} also take an interest in the adaptive sparse Transformers, which differ the selection of heads in multi-head attention, layers, and blocks in feedforward matrices between different tasks. Mixture-of-experts (MoE) is another effective approach to sparse modeling. Within the transformer model, the GShard model replaces a single feedforward (FFN) sub-layer with an MoE module consisting of multiple FFN sub-layers \citep{lepikhin21gshard}. \citet{william21switch}

\textbf{Adapters} have proven to be very efficient for NLP multi-task modeling \citep{houlsby19parameter,Bapna19simple,Pham20Study,pfeiffer20adapterhub}. The method consists of plugging several adapters to the intermediate layers of a pretrained Transformer and finetuning these adapters on the downstream tasks. Adapters can also be trained unsupervised in multilingual translation \citep{Philip20monolingual}. However, the hard-coded separation between the domains of different tasks leads to a catastrophic forgetting \citep{pfeiffer21adapterfusion}, which is a common problem in multi-task modeling using neural networks \cite{Michael89catastrophic}. In multi-domain translation, recently \citet{Pham21revisiting} demonstrated the brittleness of domain adapters against fuzzy domain separations, out-of-domain distributions, and erroneous domain tags. Recently, several works have aimed to mitigate this weakness through a mixture of expert mechanisms, for example,\cite{pfeiffer21adapterfusion}.

@@

\section{Conclusions and outlook}
In this work, we present a novel method allows jointly searching optimal sub-networks for multiple tasks and the parameters of the underlying network. Our method proposed a rigorous mathematical modeling and an end-to-end optimization with very little extra computational cost. We demonstrate an improvement of our model with large margin over multi-domain Transformer without zero additional parameters and an equivalent performance compared to a strong baseline such as Domain Adapters consisting of large number of extra parameters. In multilingual translation, our model outperforms multilingual Transformer and Language Adapters in 3/4 cases. Besides, we provided an thorough ablation study on the similarities between learned sub-networks and demonstrated a strong correlation between the sub-networks' similarity and the proximity of their tasks (domain or language).

>>>>

\section{Introduction}
Multidomain and multilingual machine translation aim to develop one single model to perform translations for multiple domains and multiple language pairs, respectively.\footnote{We will refer to these two situations as 'multi-task MT' and refer to individual domains and languages as 'tasks'.} These paradigms are motivated by the compactness of the resulting translation system \citep{Chu18multilingual,dabre20survey}, the hypothetical positive knowledge transfer between similar domains \citep{Pham21revisiting} or between languages in the same family \citep{Tan19multilingual}. However, having all the tasks use exactly the same model parameters can cause negative interference between unrelated tasks \citep{conneau20unsupervised,wang20negative}. Hence the recent development of approaches relying on a partial sharing of the parameters, eg.\ using adapter layers as studied in \citep{houlsby19parameter,Bapna19simple,Pham20Study,Philip20monolingual}. If these techniques have proven effective for building strong baselines, they fail to fully take advantage of the similarities that exist between domains and tasks, as documented eg.\ in \citep{Pham21revisiting}. This is because the partition of the parameter space between generic or task-specific subparts, and their allocation to each task is hard-coded in the network, irrespective of the actual commonalties and differences in the data space.

In this work, we develop techniques that can take the similarity between tasks into account in a more effective way, by learning the network organization from the data. To this end, we introduce a set of latent variables in the model, to account for the unseen association between tasks and regions of the parameter space, and show how training can still be performed end-to-end using a variational surrogate of the standard loss function. Our experiments with multilingual and multidomain machine translation confirm that this method can automatically detect similarities in the data, meaning that related tasks use the same subparts of the network. Our results also show that this method is comparable to using adapter layers in a number of empirical comparisons; however, contrarily to adapters, these performance are obtained without any increase of the model size. Our main contributions are summarized below:
\begin{itemize}
  \setlength{\itemsep}{1pt}
  \setlength{\parskip}{0pt}
  \setlength{\parsep}{0pt}
\item We provide a sound mathematical formulation to the problem of jointly learning task-dependent sub-networks and the parameters of the underlying models using variational probabilistic modeling techniques;
\item We present algorithms to train this model end to end with very little extra computational cost.
\item We report, using an extensive experimental setting, significant gains of this method with respect to strong baselines;
\item We study how this method can actually exploit the similarities between tasks to learn interpretable sub-networks.
\end{itemize}
\fyDone{Can you do distillation ? That is only keep the parameter for one domain and still get good results (better than DA ?)}

@@

For multilingual (TED-related) case, the data contain four language families including (1) Turkic family with Azerbaijani and Turkish(\domain{aze},\domain{tur}); (2) Slavic family with Belarusian and Russian (\domain{bel},\domain{rus}); (3) Romance family with Galician and Portuguese (\domain{glg}, \domain{por}); and (4) Czech-Slovak family with Slovak and Czech (\domain{ces}, \domain{slk}). The similarities between the sub-networks for each pair are respectively 0.95, 0.94, 0.99 and 0.96: for each language, the maximally similar one (excluding the diagonoal) is always from the same language family. 
\begin{table*}[h!]
  \centering
  \begin{tabular}{|p{1cm}|*{8}{r|}} \hline
    & \multicolumn{1}{c|}{\domain{aze}} & \multicolumn{1}{c|}{\domain{bel}}& \multicolumn{1}{c|}{\domain{ces}} & \multicolumn{1}{c|}{\domain{glg}} & \multicolumn{1}{c|}{\domain{por}} & \multicolumn{1}{c|}{\domain{rus}} &\multicolumn{1}{c|}{\domain{slk}}&\multicolumn{1}{c|}{\domain{tur}} \\ \hline 
    \domain{aze} &1.0&0.90&0.89&0.86&0.85&0.91&0.89&\SB{0.95}\\
    \domain{bel} &0.90&1.0&0.94&0.91&0.91&\SB{0.94}&0.94&0.88\\
    \domain{ces} &0.89&0.94&1.0&0.92&0.94&0.98&\SB{0.99}&0.91\\
    \domain{glg} &0.86&0.91&0.92&1.0&\SB{0.96}&0.92&0.93&0.86\\ 
    \domain{por} &0.85&0.91&0.94&\SB{0.96}&1.0&0.93&0.94&0.87\\ 
    \domain{rus} &0.91&\SB{0.94}&0.98&0.92&0.93&1.0&0.98&0.93\\
    \domain{slk} &0.89&0.94&\SB{0.99}&0.93&0.94&0.98&1.0&0.91\\
    \domain{tur} &\SB{0.95}&0.88&0.91&0.86&0.87&0.93&0.91&1.0\\
    \hline
  \end{tabular}
  \caption{Multilingual sub-networks' similarities (M2O Ted-related)}
  \label{tab:related-sim}
\end{table*}

We also plot the languages based on their dropout masks in Figure~\ref{fig:visual} using a 2d projection computed by Principal Component Analysis (PCA).
\begin{figure*}[htbp]
\begin{subfigure}{\textwidth}
  \centering
  % include first image
  \includegraphics[width=0.7\textwidth]{ted_related}  
  \caption{TED-Related}
  \label{fig:visual-related}
\end{subfigure}
\\
\begin{subfigure}{\textwidth}
  \centering
  % include second image
  \includegraphics[width=0.7\textwidth]{ted_diverse}  
  \caption{TED-Diverse}
  \label{fig:visual-diverse}
\end{subfigure}
\caption{Visualization of languages according to their dropout masks (a large vector concatenating the dropping masks of all the layers of the model) constructed by PCA.}
\label{fig:visual}
\end{figure*}

\begin{figure*}[htbp]
\includegraphics[width=0.7\textwidth]{multi_domain}
\caption{Visualization of domains according to their dropout masks (a large vector concatenating the dropping masks of all the layers of the model) constructed by PCA.}
\label{fig:domain}
\end{figure*}

\section{Related work}\label{sec:related}
Multidomain and multilingual translation systems have received considerable attention in the recent years, and a exhaustive survey is beyond the goal of this paper. Domain adaptation for neural MT is surveyed in \citep{Chu17comparison}, while multidomain MT systems are notably studied in \cite{Saunders21Asurvey,Pham21revisiting}; for multilingual MT, the reader is refered eg.\ to \citep{Chu18multilingual,dabre20survey}. We focus on the most relevant subset of this literature below.

\textbf{Language similarity} The methods developed by \citep{sen19multilingual,kong21multilingual} use language proximity to design parameter sharing strategies. The authors propose a multi-decoder model sharing the same encoder among languages and routing languages in different families to different decoders.

\textbf{Sparse Transformer} The idea of adaptive sparsity is studied in several works. For instance, \citet{xian20deep} propose to use a variable depth for different tasks. The authors aimed to match the depth of the sub-network to the complexity of the task. \citet{Gong21pay,Gong21adaptive} also take an interest in the adaptive sparse Transformers, in which differ each task triggers the selection of specific heads in multi-head attention, layers, and blocks in feedforward matrices. Mixture-of-experts (MoE) constitute another effective approach to achieve sparsity. Using the Transformer architecture, the GShard model replaces a single feedforward (FFN) sub-layer with an MoE module consisting of multiple FFN sub-layers \citep{lepikhin21gshard,william21switch} .

\textbf{Adapter modules} Adapters have proven to be very efficient for NLP multi-task modeling \citep{houlsby19parameter,Bapna19simple,Pham20Study,pfeiffer20adapterhub}. The method consists of plugging several adapters to the intermediate layers of a pretrained Transformer and finetuning these adapters on the downstream tasks. Adapters can also be trained unsupervised in multilingual translation \citep{Philip20monolingual}. However, the hard-coded separation between the domains of different tasks may lead to a catastrophic forgetting effect \citep{pfeiffer21adapterfusion}, which is a common problem in multi-task modeling using neural networks \cite{Michael89catastrophic}. In multi-domain translation, \citet{Pham21revisiting} recently demonstrated the brittleness of domain adapters against fuzzy domain separations, out-of-domain distributions, and erroneous domain tags. Several subsequent studies have aimed to mitigate this weakness through a mixture of expert mechanisms, for example,\cite{pfeiffer21adapterfusion}.

@@

\section{Conclusions and outlook}\label{sec:conclusion}
In this work, we present a novel method for multdomain and multilingual translation. It allows us to jointly search for an optimal assignment of sub-networks to tasks and learn the parameters of the underlying network. Our method relies on a sound mathematical model and an end-to-end optimization with very little extra computational cost. Experimentally, we achieve a large improvement over Transformer baseline without zero additional parameters; our performance are also comparable to that of a strong a multi-task baseline using residual adapter modules which however includes a large number of extra parameters. For multilingual translation, our model outperforms multilingual Transformer and Language Adapters in 3 our of 4 cases. Besides, we provided an thorough analysis of the similarities between learned sub-networks and demonstrated a strong correlation between the sub-networks' similarity and the proximity of the corresponding tasks (domain or language).
